---
title: "ltp2"
author: "Melissa Solem"
date: "2022-11-27"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# setup
```{r}
#loading package
library(car)

#loading data
data <- read.csv("biome-specific data for pr4 - data.csv")
attach(data)

#forming models
simple.model <- lm(pct.furious ~ 1, data = data)
complex.model <- lm(pct.furious ~ 1 + temp + precip + flood, data = data)
summary(complex.model)
```

# finding best model
```{r}
# backward stepwise selecion with AIC
step(complex.model, scope = list(lower = simple.model, upper = complex.model), direction = 'backward', k = 2)
```

# multi-collinearity check
```{r}
# did not compute as only one variable was found to be significant
```


# checking models with f-test
```{r}
c.model <- lm(pct.furious ~ 1 + temp, data = data)
s.model <- lm(pct.furious ~ 1, data = data)
anova(s.model, c.model)
summary(c.model)
```

# forming model
```{r}
# loading necessary package
library(ggplot2)

# create regression plot
ggplot(data = data,aes(temp, pct.furious)) +
  geom_point() +
  geom_smooth(method='lm', color='green3') +
  labs(x='Average Environmental Temperature (°C)', y='% with Furious Symptoms', title='Rabid Raccoons') +
  theme(plot.title = element_text(hjust=0.5, size=18, face='bold')) 
```

## models for each biome
due to the low number of observations per biome type, these models have not been vetted to a high degree.
```{r}
# splitting data by biome type
split.data <- split(data, data$biome.type)
```

desert:
```{r}
# desert

#forming models
simple.model.desert <- lm(pct.furious ~ 1, data = split.data$desert)
complex.model.desert <- lm(pct.furious ~ 1 + temp + precip, data = split.data$desert) #removed flood as variable because all locations had no flood
summary(complex.model.desert)

# backward stepwise selecion with AIC
step(complex.model.desert, scope = list(lower = simple.model.desert, upper = complex.model.desert), direction = 'backward', k = 2)

# f-test
c.model.desert <- lm(pct.furious ~ 1 + temp, data = split.data$desert)
s.model.desert <- lm(pct.furious ~ 1, data = split.data$desert)
anova(s.model.desert, c.model.desert)
summary(c.model.desert)

# desert model: pct.furious = 35.243 + 1.546(temp)
# (multiple) r-squared = 0.7383
```

temperate broadleaf forests:
```{r}
# temperate broadleaf forests

#forming models
simple.model.t.broadleaf <- lm(pct.furious ~ 1, data = split.data$t.broadleaf)
complex.model.t.broadleaf <- lm(pct.furious ~ 1 + temp + precip + flood, data = split.data$t.broadleaf) 
summary(complex.model.t.broadleaf)

# backward stepwise selecion with AIC
step(complex.model.t.broadleaf, scope = list(lower = simple.model.t.broadleaf, upper = complex.model.t.broadleaf), direction = 'backward', k = 2)

# f-test
c.model.t.broadleaf <- lm(pct.furious ~ 1 + temp, data = split.data$t.broadleaf)
s.model.t.broadleaf <- lm(pct.furious ~ 1, data = split.data$t.broadleaf)
anova(s.model.t.broadleaf, c.model.t.broadleaf)
summary(c.model.t.broadleaf)

# t.broadleaf model: pct.furious = 37.2317 + 1.3656(temp)
# (multiple) r-squared = 0.9718
```

temperate conifer forests:
```{r}
# temperate conifer forests

#forming models
simple.model.t.conifer <- lm(pct.furious ~ 1, data = split.data$t.conifer)
complex.model.t.conifer <- lm(pct.furious ~ 1 + temp + precip + flood, data = split.data$t.conifer) #removed flood as variable because all locations had no flood
summary(complex.model.t.conifer)

# backward stepwise selecion with AIC
step(complex.model.t.conifer, scope = list(lower = simple.model.t.conifer, upper = complex.model.t.conifer), direction = 'backward', k = 2)

# f-test
c.model.t.conifer <- lm(pct.furious ~ 1 + temp + precip, data = split.data$t.conifer)
s.model.t.conifer <- lm(pct.furious ~ 1, data = split.data$t.conifer)
anova(s.model.t.conifer, c.model.t.conifer)
summary(c.model.t.conifer)
#showed no significant improvement over s.model. will try 1 + temp model and see if that shows a significantly better fit

# f-test
c.model.t.conifer <- lm(pct.furious ~ 1 + temp, data = split.data$t.conifer)
s.model.t.conifer <- lm(pct.furious ~ 1, data = split.data$t.conifer)
anova(s.model.t.conifer, c.model.t.conifer)
summary(c.model.t.conifer)

# t.conifer model: pct.furious = 37.4571 + 1.3902(temp)
# (multiple) r-squared = 0.9027
```

temperate grassland
```{r}
# temperate grassland

#forming models
simple.model.t.grassland <- lm(pct.furious ~ 1, data = split.data$t.grassland)
complex.model.t.grassland <- lm(pct.furious ~ 1 + temp + precip, data = split.data$t.grassland) #removed flood as variable because all locations had no flood
summary(complex.model.t.grassland)

# backward stepwise selecion with AIC
# step(complex.model.t.grassland, scope = list(lower = simple.model.t.grassland, upper = complex.model.t.grassland), direction = 'backward', k = 2)
  # returning negative infinity AIC. more data is required to create a reliable model.

# insufficient data to create model
```

taiga
```{r}
# taiga

#forming models
simple.model.taiga <- lm(pct.furious ~ 1, data = split.data$taiga)
complex.model.taiga <- lm(pct.furious ~ 1 + temp + precip + flood, data = split.data$taiga)
summary(complex.model.taiga)

# backward stepwise selecion with AIC
step(complex.model.taiga, scope = list(lower = simple.model.taiga, upper = complex.model.taiga), direction = 'backward', k = 2)

# f-test
c.model.taiga <- lm(pct.furious ~ 1 + temp, data = split.data$taiga)
s.model.taiga <- lm(pct.furious ~ 1, data = split.data$taiga)
anova(s.model.taiga, c.model.taiga)
summary(c.model.taiga)

# taiga model: pct.furious = 39.037 + 1.795(temp)
# (multiple) r-squared = 0.426
# the f-test between pct.furious ~ 1 + temp and the intercept only model yields a p-value of 0.2325, indicating that the pct.furious ~ 1 + temp model does not fit the data significantly better than the intercept-only model. There is insufficient data collected to form a reliable model.
```

tropical broadleaf forest
```{r}
# tropical broadleaf 

#forming models
simple.model.tr.broadleaf <- lm(pct.furious ~ 1, data = split.data$tr.broadleaf )
complex.model.tr.broadleaf  <- lm(pct.furious ~ 1 + temp + precip + flood, data = split.data$tr.broadleaf)
summary(complex.model.tr.broadleaf)

# backward stepwise selection with AIC
step(complex.model.tr.broadleaf, scope = list(lower = simple.model.tr.broadleaf, upper = complex.model.tr.broadleaf), direction = 'backward', k = 2)

# f-test
c.model.tr.broadleaf <- lm(pct.furious ~ 1 + temp + precip, data = split.data$tr.broadleaf)
s.model.tr.broadleaf <- lm(pct.furious ~ 1, data = split.data$tr.broadleaf)
anova(s.model.tr.broadleaf, c.model.tr.broadleaf)
summary(c.model.tr.broadleaf)
# the complicated model did not fit the data significantly better than the intercept only model. will test pct.furious ~ 1 + temp to see if this fits data significantly better.

# f-test
c.model.tr.broadleaf <- lm(pct.furious ~ 1 + temp, data = split.data$tr.broadleaf)
s.model.tr.broadleaf <- lm(pct.furious ~ 1, data = split.data$tr.broadleaf)
anova(s.model.tr.broadleaf, c.model.tr.broadleaf)
summary(c.model.tr.broadleaf)

# the f-test between pct.furious ~ 1 + temp and the intercept only model yields a p-value of 0.1725, indicating that the pct.furious ~ 1 + temp model does not fit the data significantly better than the intercept-only model. There is insufficient data collected to form a reliable model.
```

tropical conifer
```{r}
# tropical conifer

#forming models
simple.model.t.conifer <- lm(pct.furious ~ 1, data = split.data$t.conifer)
complex.model.t.conifer <- lm(pct.furious ~ 1 + temp + precip + flood, data = split.data$t.conifer)
summary(complex.model.t.conifer)

# backward stepwise selecion with AIC
step(complex.model.t.conifer, scope = list(lower = simple.model.t.conifer, upper = complex.model.t.conifer), direction = 'backward', k = 2)

# collinearity check
cc.model <- lm(pct.furious ~ 1 + temp + precip, data = split.data$t.conifer)
pairs(pct.furious ~ 1 + temp + precip, data = split.data$t.conifer)
vif(cc.model)
# no collinearity observed

# f-test
c.model.t.conifer <- lm(pct.furious ~ 1 + temp + precip, data = split.data$t.conifer)
s.model.t.conifer <- lm(pct.furious ~ 1, data = split.data$t.conifer)
anova(s.model.t.conifer, c.model.t.conifer)
summary(c.model.t.conifer)

# although the p-value of the f-test (0.05204) was above the significance level, it was near significant and will thus be included in the report.
# t.conifer model: pct.furious = 39.18569 + 1.69875(temp) - 0.06962(precip)
# (multiple) r-squared = 0.948
```

tundra
```{r}
# tundra

#forming models
simple.model.tundra <- lm(pct.furious ~ 1, data = split.data$tundra)
complex.model.tundra <- lm(pct.furious ~ 1 + temp + precip, data = split.data$tundra) #flood variable removed because no floods were observed in any observation
summary(complex.model.tundra)

# backward stepwise selecion with AIC
step(complex.model.tundra, scope = list(lower = simple.model.tundra, upper = complex.model.tundra), direction = 'backward', k = 2)

# f-test
c.model.tundra <- lm(pct.furious ~ 1 + precip, data = split.data$tundra)
s.model.tundra <- lm(pct.furious ~ 1, data = split.data$tundra)
anova(s.model.tundra, c.model.tundra)
summary(c.model.tundra)

# tundra model: pct.furious = 25.85542 + 0.24069(precip)
# (multiple) r-squared = 0.9713
```

# creating model with all plots

tropical conifer can't be plotted
```{r}
# create regression plot
ggplot(data = data,aes(temp, pct.furious, color = biome.type)) +
  geom_point() +
  scale_color_manual(name = 'Biome', 
    breaks = c('General', 'desert', 't.broadleaf', 't.conifer', 'tr.broadleaf'), 
    values = c('General' = 'black', "desert" = 'springgreen4', "t.broadleaf" = 'palegreen3', "t.conifer" = 'dodgerblue2', "tr.broadleaf" = 'skyblue3')) +
  geom_smooth(method='lm', aes(color='General')) +
  geom_smooth(data = split.data$desert, method='lm', aes(color='desert')) +
  geom_smooth(data = split.data$t.broadleaf, method='lm', aes(color='t.broadleaf')) +
  geom_smooth(data = split.data$t.conifer, method='lm', aes(color='t.conifer')) +
  geom_smooth(data = split.data$tr.broadleaf, method='lm', aes(color='tr.broadleaf')) +
  labs(x='Average Environmental Temperature (degrees C)', y='% with Furious Symptoms', title='Rabid Raccoons') +
  theme(plot.title = element_text(hjust=0.5, size=18, face='bold'))
```


# Tundra plot
```{r}
# create regression plot
ggplot(data = split.data$tundra,aes(precip, pct.furious)) +
  geom_point() +
  geom_smooth(method='lm', color='blue') +
  labs(x='Average Monthly Precipitation (mm)', y='% with Furious Symptoms', title='Rabid Raccoons in Tundra Biome') +
  theme(plot.title = element_text(hjust=0.5, size=18, face='bold'))
```

